---
title: "main_analysis"
authors: evan, laure
format: html
editor: visual
---

# Portal Project Rodent Study

## 1. Setup and Data Import

Load the tidyverse package we need for data manipulation and visualization.

```{r}
library(tidyverse)
```

Read in the two CSV files from the data folder, and take a look at each table.

```{r}
surveys <- read_csv("../data/surveys.csv")
species <- read_csv("../data/species.csv")

glimpse(surveys)
glimpse(species)
```

## 2. Data Tidying and Cleaning

Clean the surveys table by removing rows with no species ID, combining the separate year, month, and day columns into a single date, and converting sex and species ID to categorical variables.

```{r}
surveys_clean <- surveys |>
  filter(!is.na(species_id)) |>
  mutate(date = make_date(year, month, day)) |>
  mutate(
    sex        = factor(sex, levels = c("M", "F")),
    species_id = factor(species_id)
  )
```

Count how many rows are missing values in key columns so we know what we are working with before filtering.

```{r}
surveys_clean |>
  summarise(
    missing_sex             = sum(is.na(sex)),
    missing_hindfoot_length = sum(is.na(hindfoot_length)),
    missing_weight          = sum(is.na(weight))
  )
```

## 3. Joining Tables

Join the cleaned surveys table to the species table so each observation has the full species name, genus, and taxa group attached to it, then confirm the join worked and the new columns are present

```{r}
surveys_joined <- surveys_clean |>
  left_join(species, by = "species_id")

glimpse(surveys_joined)
```

## 4. Filter to Rodents and Handle Missing Values

Filter down to only rodents since they make up the vast majority of observations and are the main focus of this dataset, then check how many rows we have.

```{r}
rodents <- surveys_joined |>
  filter(taxa == "Rodent")

nrow(rodents)
```

Remove rows missing weight, hindfoot length, or sex since these are key variables we could analyze, then check how many rows remain after cleaning.

```{r}
rodents_clean <- rodents |>
  filter(
    !is.na(weight),
    !is.na(hindfoot_length),
    !is.na(sex)
  )

nrow(rodents_clean)
```

## 5. Data Transformation and Species Labels

Calculate the mean weight and average hindfoot length for each species, then rank them by how commonly they appear in the dataset.

```{r}
species_summary <- rodents_clean |>
  group_by(species_id, genus, species) |>
  summarise(
    mean_weight    = mean(weight),
    mean_hindfoot  = mean(hindfoot_length),
    count          = n(),
    .groups        = "drop"
  ) |>
  arrange(desc(count))

species_summary
```

Create a clean species label by combining genus and species name into a string, and select only the columns we need for analysis. Check if the new column looks correct.

```{r}
rodents_clean <- rodents_clean |>
  mutate(species_label = str_c(genus, species, sep = " "))

rodents_clean |>
  select(genus, species, species_label) |>
  distinct() |>
  arrange(species_label)

view(rodents_clean)
```

Identify the top 5 most common species to keep the plot readable.

```{r}
top5_species <- rodents_clean |>
  count(species_label, sort = TRUE) |>
  slice_head(n = 5) |>
  pull(species_label)

glimpse(top5_species)
```

## 6. Visualization 1 - Weight Distribution by Species

Plot weight distributions for the top 5 species using a density plot.

Density shows the probability of observing a rodent at a given weight. Peaks show where most individuals cluster, wider curves mean more variation.

```{r}
rodents_clean |>
  filter(species_label %in% top5_species) |>
  mutate(species_label = fct_reorder(species_label, weight, median)) |>
  ggplot(aes(x = weight, fill = species_label, color = species_label)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Body Weight Distribution By Species",
    x     = "Weight (g)",
    y     = "Density",
    fill  = "Species",
    color = "Species"
  )
```

## 7. Visualization 2 - Mean Weight Over Time by Species

Calculate mean weight per year and trends over time to see how rodent body weight has changed across the study period.

```{r}
rodents_clean |>
  group_by(year, species_label) |>
  summarise(
    mean_weight = mean(weight),
    .groups     = "drop"
  ) |>
  filter(species_label %in% top5_species) |>
  ggplot(aes(x = year, y = mean_weight, color = species_label)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mean Rodent Body Weight Over Time by Species",
    x     = "Year",
    y     = "Mean Weight (g)",
    color = "Species"
  )
```

## 8. Visualization 3 - Mean Weight by Species and Sex

Compare mean body weight between male and female rodents across the top 5 most common species to see whether weight differs by sex.

```{r}
rodents_clean |>
  filter(species_label %in% top5_species) |>
  group_by(species_label, sex) |>
  summarise(
    mean_weight = mean(weight),
    .groups     = "drop"
  ) |>
  mutate(species_label = fct_reorder(species_label, mean_weight, max)) |>
  ggplot(aes(x = species_label, y = mean_weight, fill = sex)) +
  geom_col(position = "dodge") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(
    title = "Mean Body Weight by Species and Sex",
    x     = "Species",
    y     = "Mean Weight (g)",
    fill  = "Sex"
  )
```
