---
title: "main_analysis"
authors: evan, laure
format: html
editor: visual
---

## 1. Setup and Data Import

Load the packages we need for data manipulation and visualization.

```{r}
library(tidyverse)
library(readr)
library(ggplot2)
```

Read in the two CSV files from the data folder, and take a look at each table.

```{r}
surveys <- read_csv("data/surveys.csv")
species <- read_csv("data/species.csv")

glimpse(surveys)
glimpse(species)
```

## 2. Data Tidying and Cleaning

Clean the surveys table by removing rows with no species ID, combining the separate year, month, and day columns into a single date, and converting sex and species ID to categorical variables.

```{r}
surveys_clean <- surveys |>
  filter(!is.na(species_id)) |>
  mutate(date = make_date(year, month, day)) |>
  mutate(
    sex        = factor(sex, levels = c("M", "F")),
    species_id = factor(species_id)
  )
```

Count how many rows are missing values in key columns so we know what we are working with before filtering.

```{r}
surveys_clean |>
  summarise(
    missing_sex             = sum(is.na(sex)),
    missing_hindfoot_length = sum(is.na(hindfoot_length)),
    missing_weight          = sum(is.na(weight))
  )
```

## 3. Joining Tables

Join the cleaned surveys table to the species table so each observation has the full species name, genus, and taxa group attached to it, then confirm the join worked and the new columns are present

```{r}
surveys_joined <- surveys_clean |>
  left_join(species, by = "species_id")

glimpse(surveys_joined)
```

## 4. Filter to Rodents and Handle Missing Values

Filter down to only rodents since they make up the vast majority of observations and are the main focus of this dataset, then check how many rows we have.

```{r}
rodents <- surveys_joined |>
  filter(taxa == "Rodent")

nrow(rodents)
```

Remove rows missing weight, hindfoot length, or sex since these are the key variables we will be analyzing, then check how many rows remain after cleaning.

```{r}
rodents_clean <- rodents |>
  filter(
    !is.na(weight),
    !is.na(hindfoot_length),
    !is.na(sex)
  )

nrow(rodents_clean)
```

## 5. Data Transformation and Species Labels

Calculate the mean weight and average hindfoot length for each species, then rank them by how commonly they appear in the dataset.

```{r}
species_summary <- rodents_clean |>
  group_by(species_id, genus, species) |>
  summarise(
    mean_weight    = mean(weight),
    mean_hindfoot  = mean(hindfoot_length),
    count          = n(),
    .groups        = "drop"
  ) |>
  arrange(desc(count))

species_summary
```

Create a clean species label by combining genus and species name into a string, and select only the columns we need for analysis. Check if the new column looks correct.

```{r}
rodents_clean <- rodents_clean |>
  mutate(species_label = str_c(genus, species, sep = " "))

rodents_clean |>
  select(genus, species, species_label) |>
  distinct() |>
  arrange(species_label)
```

Identify the top 5 most common species to keep the plot readable.

```{r}
top5_species <- rodents_clean |>
  count(species_label, sort = TRUE) |>
  slice_head(n = 5) |>
  pull(species_label)

glimpse(top5_species)
```
